---
title: "week3"
author: "Emin Allen"
date: "2024-09-21"
output: html_document
---


```{r, include=FALSE}
# Supress package startup messages
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(knitr))
```

```{r}
# Load libraries
library(dplyr)
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

In Week 3, we work with data normalization and character manipulation. In this assignment, Part 1 will be normalizing data and Part 2 will involve working with a College Majors dataset from the fivethirtyeight website.   

# *Part 1: Normalization*

Normalization in data science means organizing data in a way that reduces redundancy and improves data integrity. This process involves breaking down larger tables into smaller tables. Here are three dataframes to illustrate the normalization process:

## Example 1: Normalizating Employee Data

The original employee dataframe contains information about employees, departments, and salaries. 

### Create original Employee dataframe
```{r}
employee_df <- data.frame(
  employee_id = c(1, 2, 3, 4),
  name = c("Gary Larsen", "William Pace", "Jennifer Fisher", "Samuel Morales"),
  department = c("Finance", "IT", "Finance", "HR"),
  salary = c(95000, 85000, 120000, 65000)
)
```

### Display original Employee dataframe
```{r}
knitr::kable(employee_df)
```

We normalize the original Employee dataframe into separate Employees, Departments, and Salaries tables. 

### Create normalized Employees table
```{r}
employees_df <- data.frame(
  employee_id = c(1, 2, 3, 4),
  name = c("Gary Larsen", "William Pace", "Jennifer Fisher", "Samuel Morales")
)
```

### Display normalized Employees table
```{r}
knitr::kable(employees_df)
```

### Create normalized Departments table
```{r}
departments_df <- data.frame(
  department_id = c(1,2,3),
  department = c("Finance", "IT", "HR")
)
```

### Display normalized Departments table
```{r}
knitr::kable(departments_df)
```

### Create normalized Salaries table
```{r}
salaries_df <- data.frame(
  employee_id = c(1, 2, 3, 4),
  department_id = c(1, 2, 1, 3),
  salary = c(95000, 85000, 120000, 65000)
)
```
### Display normalized Salaries table

```{r}
knitr::kable(salaries_df)
```

# Example 2: Normalizing Product Inventory Data

The original product inventory dataframe contains information such as the product ID, name, category, quantity, and price. 

### Create original Product Inventory dataframe
```{r}
inventory_df <- data.frame(
  product_id = c(1,2,3,4),
  product_name = c("Keyboard", "Dumbbell", "Running Shoes", "Water Bottle"),
  category = c("Computer Accessories", "Fitness Equipment", "Sportswear", "Hydration"),
  quantity = c(50, 30, 20, 100),
  price = c(40, 100, 150, 20)
)
```


### Display original Product Inventory dataframe
```{r}
knitr::kable(inventory_df)
```

We normalize the original Product Inventory dataframe into separate Products, Categories, and Inventory tables. 

### Create normalized Products table
```{r}
products_df <- data.frame(
  product_id = c(1, 2, 3, 4), 
  product_name = c("Keyboard", "Dumbbell", "Running Shoes", "Water Bottle")
)
```

### Display normalized Products table
```{r}
knitr::kable(products_df)
```

### Create normalized Categories table
```{r}
categories_df <- data.frame(
  category_id = c(1, 2, 3, 4),
  category = c("Computer Accessories", "Fitness Equipment", "Sportswear", "Hydration")
)
```

### Display normalized Categories table
```{r}
knitr::kable(categories_df)
```

### Create normalized Inventory table
```{r}
inventory_details_df <- data.frame(
  product_id = c(1, 2, 3, 4),
  category_id = c(1, 2, 3, 4),
  quantity = c(50, 30, 20, 100),
  price = c(40, 100, 150, 20)
)
```

### Display normalized Inventory table
```{r}
knitr::kable(inventory_details_df)
```

# Example 3: Normalizing Sales Data

The original Sales dataframe contains information such as the Sale ID, Product name, Customer, and Sale amount.

### Create original Sales dataframe
```{r}
sales_df <- data.frame(
  sale_id = c(1, 2, 3, 4),
  product_name = c("Soda", "Bagel", "Coffee", "Cheese"), 
  customer = c("James", "Shirley", "Mike", "Roman"),
  sale_amount = c(2.50, 4.50, 1.50, 6.00)
)
```

### Display original Sales dataframe
```{r}
knitr::kable(sales_df)
```

### Create normalized Products table
```{r}
products_sales_df <- data.frame(
  product_id = c(1, 2, 3, 4), 
  product_name = c("Soda", "Bagel", "Coffee", "Cheese")
)
```

### Display normalized Products table

```{r}
knitr::kable(products_sales_df)
```

### Create normalized Customers table
```{r}
customers_df <- data.frame(
  customer_id = c(1, 2, 3, 4),
  customer = c("James", "Shirley", "Mike", "Roman")
)
```

### Display normalized Customers table
knitr::kable(customers_df)

### Create normalized Sales table
```{r}
sales_details_df <- data.frame(
  sale_id = c(1, 2, 3, 4),
  product_id = c(1, 2, 3, 4),
  customer_id = c(1, 2, 3, 4),
  sale_amount = c(2.50, 4.50, 1.50, 6.00)
)
```


### Display normalized Sales table
```{r}
knitr::kable(sales_details_df)
```

# *Part 2: Character Manipulation*
Source: [https://fivethirtyeight.com/features/the-economic-guide-to-picking-a-college-major/]

Using the 173 majors listed in fivethirtyeight College Majors dataset, provide code that identifies the majors that contain either "DATA" or "STATISTICS"

### Load the College Majors dataset
```{r}
majors_df <- read.csv("https://raw.githubusercontent.com/fivethirtyeight/data/refs/heads/master/college-majors/all-ages.csv")
```

### Find majors containing "DATA" or "STATISTICS"
```{r}
selected_majors <- majors_df[grep("DATA|STATITICS", majors_df$Major, ignore.case = TRUE), ]
```

### Display selected majors
```{r}
knitr::kable(selected_majors)
```

### Clean selected majors for better readability
```{r}
cleaned_majors <- selected_majors %>%
  select(Major, Employed, Unemployment_rate, Median) %>%
  rename(
    Major_Name = Major,
    Employed_Graduates = Employed,
    Unemployment_Rate = Unemployment_rate,
    Median_Salary = Median
  )
```

```{r}
knitr::kable(cleaned_majors)
```



# *Part 2: Regular Expressions*
Describe, in words, what these expressions will match:


`(.)\1\1`

**Description**: Looks for the same character repeated 3x in a row

**Example**: "aaa", "111", "!!!"

`"(.)(.)\\2\\1"`

**Description**: Looks for two different characters next to each other, followed by those same two characters but mirrored in reverse (like a symmetrical reflection). 

**Example**: "abba", "7887"


`(..)\1`

**Description**: Looks for two characters repeated one after the other.

**Example**: "abab", "7878"



`"(.).\\1.\\1"`

**Description**: Matches single character, any character, first character, any character, first character. 

**Example**: "abaca", "12131"


`"(.)(.)(.).*\\3\\2\\1"`

**Description**: Looks for 3 characters, followed by any sequence of characters, an then those 3 characters in reverse. 

**Example**: "abc...cba", "123...321"


# Construct regular expressions to match words that:

**Question 1**: Start and end with the same character.

**Regex Answer**: ^(.).*\1$


**Question 2**: Contain a repeated pair of letters (e.g. "church contains "ch" repeated twice)

**Regex Answer**: (..).*\1

**Question 3**: Contain one letter repeated in at least three places (e.g. "eleven" contains three "e"s)

**Regex Answer**: (.).*?\1.*?\1

